### 2021-09-29 documents/sections/verbs.tex
working on overview of etxensions and verbs
work on verbs section, added verb cognates

\subsection{\rc{(ət-)jəpɨ} \qu{to come}}
\label{sec:come}
\definecolor{come1}{RGB}{161.0,201.0,244.0}
\definecolor{come2}{RGB}{255.0,180.0,130.0}
\definecolor{come3}{RGB}{141.0,229.0,161.0}
\definecolor{come4}{RGB}{255.0,159.0,155.0}
\begin{table}
	\centering
	\caption{Reflexes of \qu{to come}}
    \label{tab:come}
	\begin{tabular}{@{}lll@{}}
	\mytoprule
	Language & Form & Source\\
	\mymidrule
\kaxui & \colorbox{come3}{\obj{ehɨ}} & \perscomm{Spike Gildea}\\
\kaxui & \colorbox{come1}{\obj{johɨ}} & \perscomm{Spike Gildea}\\
\kaxui & \colorbox{come2}{\obj{oohɨ}} & \perscomm{Spike Gildea}\\
\arara & \colorbox{come3}{\obj{ebɨ}} & \textcite[150]{alves2017arara}\\
\arara & \colorbox{come4}{\obj{odebɨ}} & \textcite[150]{alves2017arara}\\
\ikpeng & \colorbox{come4}{\obj{arep}} & \textcite[265]{ikpengpacheco2001}\\
\bakairi & \colorbox{come4}{\obj{əewɨ}} & \textcite[4]{meira2003bakairi}\\
\trio & \colorbox{come4}{\obj{əepɨ}} & \textcite[294]{triomeira1999}\\
\trio & \colorbox{come3}{\obj{epɨ}} & \textcite[294]{triomeira1999}\\
\akuriyo & \colorbox{come3}{\obj{epɨ}} & \textcite[168]{meira1998proto}\\
\carijo & \colorbox{come3}{\obj{ehɨ}} & \textcite[178]{robayo2000avance}\\
\apalai & \colorbox{come4}{\obj{oepɨ}} & \textcite[37]{koehn1986apalai}\\
\kalina & \colorbox{come2}{\obj{opɨ}} & \textcite[429]{courtz2008carib}\\
\maqui & \colorbox{come3}{\obj{ehə}} & \textcite[438]{maquiritaricaceres2011}\\
\akawaio & \colorbox{come4}{\obj{əsipɨ}} & \textcite[160]{stegeman2014akawaio}\\
\akawaio & \colorbox{come1}{\obj{jepɨ}} & \textcite[125]{akawaiocaesar2003}\\
\ingariko & \colorbox{come1}{\obj{jepə}} & \textcite[415]{cruz2005fonologia}\\
\ingariko & \colorbox{come1}{\obj{jə}} & \textcite[299]{cruz2005fonologia}\\
\patamona & \colorbox{come1}{\obj{jepɨ}} & \perscomm{Spike Gildea}\\
\patamona & \colorbox{come1}{\obj{jəpɨ}} & \perscomm{Spike Gildea}\\
\pemon & \colorbox{come1}{\obj{jepɨ}} & \textcite[102]{alvarez2000construcciones}\\
\macushi & \colorbox{come1}{\obj{ipɨ}} & \textcite[32]{macushiabbott1991}\\
\panare & \colorbox{come2}{\obj{əpɨ}} & \textcite[65]{panarepayne2013}\\
\yawarana & \colorbox{come2}{\obj{əpɨ}} & \textcite[68]{mendez1959yawarana}\\
\mapoyo & \colorbox{come3}{\obj{epɨ}} & \textcite[74]{muller1975mapoyo}\\
\uxc & \colorbox{come2}{\obj{ee}} & \textcite[182]{meira2005southern}\\
\mybottomrule
	\end{tabular}
\end{table}
This verb is reconstructed as monomorphemic \rc{ətepɨ} by \textcite[30]{gildea2007greenberg}, but an inspection of all the attested reflexes \pcref{tab:come} suggests a more complex story.
In particular, the Pemongan languages and one of its allomorphs in \kaxui suggest a root \colorbox{come1}{\rc{jəpɨ}}.
However, the one form in \kaxui only occurs in the Progressive \exref{kaxprog.kax-82}, and \obj{j(-)} may be a reflex the \settwo third person marker \rc{i-} combined with  the \gl{s_a_} class marker \rc{w-}, as suggested by the occurrence of the latter with other person values \exref{kaxprog.kax-54}.
On the other hand, while C-initial verbs do show a clear reflex of third person \rc{i-} \exref{kaxprog.kax-77}, regular V-initial verbs do not show \obj{j-}, but Ø \exref{kaxprog.kax-83}.
Thus, it seems more likely that the \obj{j} is indeed part of the verb, rather than an outcome of \rc{i-w-}.

@@

There is further evidence supporting the reconstruction of both \rc{j} and \rc{ə}.
First, there are many forms reflecting \colorbox{come2}{\rc{əpɨ}}, occurring in different branches of the family.\footnote{\obj{e} is the regular outcome of \rc{ə} in \uxc; \rc{e} became \obj{i} \parencite[176]{meira2005southern}.}
Second, there are also forms reflecting \colorbox{come3}{\rc{epɨ}}, and umlaut of \rc{ə} to \rc{e} is conditioned by preceding \rc{j} \parencite{meira2010origin}.
That is, a unifying account of these forms starts from a root \rc{jəpɨ}, which is the subject to two major sound changes: \begin{inlinelist}
 \item \rc{j}-loss
 \item \rc{ə}-umlaut	
 \end{inlinelist}.
However, these sound changes appear to have applied irregularly, and not always in the same order.
For example, the \kalina form \obj{opɨ} can only be explained if \rc{j} was lost before the umlaut of \rc{ə} to \rc{e}, which happened elsewhere after \rc{j}.
On the other hand, forms like \maqui \obj{ehə} must be the result of \phonl{\rc{ə}}{\rc{e}}{\rc{j}}, with subsequent loss of \rc{j}.

we have so far ignored the forms which form the basis for \posscite{gildea2007greenberg} \rc{ətepɨ}; these are compatible with our reconstructed root \rc{jəpɨ}, with the addition of the detransitivizer \rc{ət(e)-}: \colorbox{come4}{\rc{ətjəpɨ}}.
In fact, the \obj{i} in the \akawaio form \obj{əsipɨ} is better explained by a sequence \rc{jə}, which has the same outcome \obj{i} in the \macushi reflex of \rc{jəpɨ}, \obj{ipɨ}.
While the semantics of combining a detransitivizer with an intransitive verb do not really make sense, some historical \gl{s_p_} verbs are attested as adding the detransitivizer to become \gl{s_a_} verbs.
For example, \gl{s_p_} \rc{wənɨkɨ} \qu{to sleep} becomes \trio \obj{əənɨkɨ} \parencite[252]{triomeira1999} and \kalina \obj{əʔnɨkɨ} \parencite[429]{courtz2008carib}, both \gl{s_a_}.
\waiwai \qu{go to sleep} can be \obj{wɨnɨk} \parencite[30]{waiwaihawkins1998} or \obj{et-wɨnɨk} \parencite[204]{hawkins1953waiwai}.
The parallels to \qu{to sleep} end here, since bare \rc{jəpɨ} \qu{to come} apparently already was an \gl{s_a_} verb, as evidenced by its status in \kaxui, \kalina, \panare \exref{pan-128}, \arara, \trio, and \akuriyo.

@@

While these sound changes and the addition of \rc{ət(e)-} do account for the majority of the forms in \cref{tab:come},\footnote{The only exception is the \apalai form \obj{oepɨ}, where the detransitivizer would have the reflex \obj{os-} \parencite[506]{meira2010origin}. While \obj{oepɨ} would be a regular outcome of \rc{ə-jəpɨ}, the \envr{}{C} allomorph of the detransitivizer is \obj{e-}. The form may be due to borrowing from \trio, which has lost intervocalic \rc{t} to create \obj{əepɨ}. Alternatively, \apalai \obj{oepɨ} could be a fossilized loan \wayana, which has lost its reflex of \rc{ətjəpɨ}, but where regular sound changes would also have resulted in the loss of intervocalic \rc{t} \parencite[63]{wayanatavares2005}.}
the distribution within the family is rather chaotic.
Not only do very closely related languages show different forms, like \yawarana and \mapoyo, but distinct forms can even be found within the same language, usually conditioned by different prefixes.
This was discussed in \cref{sec:pekodian} for \arara, which has reflexes of \rc{ətjəpɨ} and \rc{jəpɨ} within the same paradigm.
A similar situation is found in \trio, where the \setone paradigm shows a reflex of \rc{ətjəpɨ} for first, but of \rc{epɨ} (< \rc{jəpɨ}) for the other persons \exref{tricome}.\footnote{While the \gl{1+2} form is a regular outcome of \rc{kɨt-epɨ}, the second person form is mysterious.}

\ex<tricome> \trio \parencite[294]{triomeira1999}\\
\begin{tabular}[t]{@{}ll@{}}
\gl{1} & \obj{w-\colorbox{come4}{əepɨ}} \\
\gl{2} &  \obj{mən-\colorbox{come3}{epɨ}} \\ 
\gl{1+2} &  \obj{ke-\colorbox{come3}{epɨ}} \\
\gl{3} &  \obj{n-\colorbox{come3}{epɨ}} \\
\end{tabular}
\xe
%
Summing up, this verb is highly irregular, both from a synchronic and diachronic perspective.
It seems that reflexes of the detransitivizer \rc{ət(e)-} were optionally added to an \gl{s_a_} verb root \rc{jəpɨ}, which further underwent umlaut and loss of \rc{j}, but in no systematic manner, resulting in the chaotic picture in \cref{tab:come}.

As discussed in \cref{sec:pekodian}, innovative \rc{k-} was introduced on the \arara reflex of \rc{jəpɨ}, but not on the \ikpeng and \bakairi reflexes of \rc{ətjəpɨ}.
Both reflexes of \rc{ətjəpɨ} (\trio) and of \rc{jəpɨ} (\akuriyo) resisted the introduction of \rc{t-} in \PTir.
\carijo \obj{ehɨ} shows innovative \obj{j-}, rather than conservative \obj{w-} \exref{car-18}.
It is unknown whether there is a \yukpa reflex of this verb, and it was fully replaced in \PWai by \rc{omokɨ} \qu{to come}.

@@

>>>>

\subsection{\rc{(ət-)jəpɨ} \qu{to come}}
\label{sec:come}
\begin{table}
	\centering
	\caption{Reflexes of \qu{to come} \parencites[150]{alves2017arara}[265]{ikpengpacheco2001}[4]{meira2003bakairi}[294]{triomeira1999}[168]{meira1998proto}[178]{robayo2000avance}[37]{koehn1986apalai}[429]{courtz2008carib}[438]{maquiritaricaceres2011}[160]{stegeman2014akawaio}[125]{akawaiocaesar2003}[299, 415]{cruz2005fonologia}[102]{alvarez2000construcciones}[32]{macushiabbott1991}[65]{panarepayne2013}[68]{mendez1959yawarana}[74]{muller1975mapoyo}[182; p.c., Spike Gildea]{meira2005southern}}
    \label{tab:come}
\begin{tabular}[t]{@{}ll@{}}
\toprule
Language &                               Form \\
\midrule
\kaxui    &  \obj{johɨ}, \obj{ehɨ}, \obj{oohɨ} \\
\arara    &             \obj{ebɨ}, \obj{odebɨ} \\
\ikpeng   &                         \obj{arep} \\
\bakairi  &                         \obj{əewɨ} \\
\trio     &              \obj{epɨ}, \obj{əepɨ} \\
\akuriyo  &                         \obj{eepɨ} \\
\carijo   &                          \obj{ehɨ} \\
\apalai   &                         \obj{oepɨ} \\
\kalina   &                          \obj{opɨ} \\
\maqui    &                          \obj{ehə} \\
\akawaio  &            \obj{jepɨ}, \obj{əsipɨ} \\
\ingariko &               \obj{jepə}, \obj{jə} \\
\patamona &             \obj{jəpɨ}, \obj{jepɨ} \\
\pemon    &                         \obj{jepɨ} \\
\macushi  &                          \obj{ipɨ} \\
\panare   &                          \obj{əpɨ} \\
\yawarana &                          \obj{əpɨ} \\
\mapoyo   &                          \obj{epɨ} \\
\uxc      &                           \obj{ee} \\
\bottomrule
\end{tabular}
\end{table}

This verb is reconstructed as monomorphemic \rc{ətepɨ} by \textcite[30]{gildea2007greenberg}, but an inspection of all the attested reflexes \pcref{tab:come} suggests a more complex story.
Different languages present rather different forms, which we argue are all (partially) cognate, with the reconstructed \PC verb having the form \rc{(ət-)jəpɨ}.
Reflexes of a bare \rc{j-}initial root \rc{jəpɨ} are found in the Pemongan languages and \kaxui.
However, the one form in \kaxui only occurs in the Progressive \exref{kaxprog.kax-82}, and \obj{j(-)} may be a reflex the \settwo third person marker \rc{i-} combined with  the \gl{s_a_} class marker \rc{w-}, as suggested by the occurrence of the latter with other person values \exref{kaxprog.kax-54}.
On the other hand, while C-initial verbs do show a clear reflex of third person \rc{i-} \exref{kaxprog.kax-77}, regular V-initial verbs do not show \obj{j-}, but Ø \exref{kaxprog.kax-83}.
Thus, it seems more likely that the \obj{j} is indeed part of the root, rather than an outcome of \rc{i-w-}.

@@

Other languages -- across the family -- have forms reflecting the same root without the initial consonant, \rc{əpɨ}: \kalina, \panare, \yawarana, and \uxc.\footnote{\obj{e} is the regular outcome of \rc{ə} in \uxc; \rc{e} became \obj{i} \parencite[176]{meira2005southern}.}
There are also forms reflecting \rc{epɨ}, again across the family: \kaxui, \arara, \trio, \carijo, \maqui, and \mapoyo.
The \akuriyo form \obj{eepɨ} looks like a reflex thereof as well, but the length is unexpected, and is analyzed by \textcite[]{meira1998proto} as reflecting an earlier diphthong \rc{əe}.
A unifying account of these forms starts from a root \rc{jəpɨ}, which is the subject to two major sound changes: \begin{inlinelist}
 \item \rc{j}-loss
 \item \rc{ə}-umlaut after \rc{j}
 \end{inlinelist}.
Both phenomena are widespread in the family \parencite{meira2010origin}.
However, these sound changes appear to have applied irregularly, and not always in the same order.
For example, the \kalina form \obj{opɨ} can only be explained if \rc{j} was lost before the umlaut of \rc{ə} to \rc{e}, which happened elsewhere after \rc{j}.
On the other hand, forms like \maqui \obj{ehə} must be the result of \phonl{\rc{ə}}{\rc{e}}{\rc{j}}, with subsequent loss of \rc{j}.

We have so far ignored the forms which form the basis for \posscite{gildea2007greenberg} \rc{ətepɨ}; these are compatible with our reconstructed root \rc{jəpɨ}, with the addition of the detransitivizer \rc{ət(e)-}: \rc{ətjəpɨ}.
While most of the modern forms are lacking evidence for \rc{j} and thus look like a reflex of \rc{ətepɨ}, the \obj{i} in the \akawaio form \obj{əsipɨ} is evidence of the sequence \rc{jə}, which has the same outcome \obj{i} in the \macushi reflex of \rc{jəpɨ}, \obj{ipɨ}.
Although the semantics of combining a detransitivizer with an intransitive verb do not really make sense, some historical \gl{s_p_} verbs are attested as adding the detransitivizer to become \gl{s_a_} verbs.
For example, \gl{s_p_} \rc{wənɨkɨ} \qu{to sleep} becomes \trio \obj{əənɨkɨ} \parencite[252]{triomeira1999} and \kalina \obj{əʔnɨkɨ} \parencite[429]{courtz2008carib}, both \gl{s_a_}.
\waiwai \qu{go to sleep} can be \obj{wɨnɨk} \parencite[30]{waiwaihawkins1998} or \obj{et-wɨnɨk} \parencite[204]{hawkins1953waiwai}.
The parallels to \qu{to sleep} end here, since bare \rc{jəpɨ} \qu{to come} apparently already was an \gl{s_a_} verb, as evidenced by its status in \kaxui, \kalina, \panare \exref{pan-128}, \arara, \trio, and \akuriyo.

@@

While these sound changes and the addition of \rc{ət(e)-} do account for the majority of the forms in \cref{tab:come},\footnote{
Apart from aforementioned \akuriyo \obj{eepɨ}, another exception is the \apalai form \obj{oepɨ}, where the detransitivizer would have the reflex \obj{os-} \parencite[506]{meira2010origin}.
While \obj{oepɨ} would be a regular outcome of \rc{ə-jəpɨ}, the \envr{}{C} allomorph of the detransitivizer is \obj{e-}.
The form may be due to borrowing from \trio, which has lost intervocalic \rc{t} to create \obj{əepɨ}.
Alternatively, \apalai \obj{oepɨ} could be a fossilized loan \wayana, which has lost its reflex of \rc{ətjəpɨ}, but where regular sound changes would also have resulted in the loss of intervocalic \rc{t} \parencite[63]{wayanatavares2005}.}
the distribution within the family is rather chaotic.
Not only do very closely related languages show different forms, like \yawarana and \mapoyo, but distinct forms can even be found within the same language, usually conditioned by different prefixes.
This was discussed in \cref{sec:pekodian} for \arara, which has reflexes of \rc{ətjəpɨ} and \rc{jəpɨ} within the same paradigm.
A similar situation is found in \trio, where the \setone paradigm shows a reflex of \rc{ətepɨ} (< \rc{ətjəpɨ}) for first, but of \rc{epɨ} (< \rc{jəpɨ}) for the other persons \exref{tricome}.\footnote{While the \gl{1+2} form is a regular outcome of \rc{kɨt-epɨ}, the second person form is mysterious.}

\ex<tricome> \trio \parencite[294]{triomeira1999}\\
\begin{tabular}[t]{@{}ll@{}}
\gl{1} & \obj{w-əepɨ} \\
\gl{2} &  \obj{mən-epɨ} \\ 
\gl{1+2} &  \obj{ke-epɨ} \\
\gl{3} &  \obj{n-epɨ} \\
\end{tabular}
\xe
%
Summing up, this verb is highly irregular, both from a synchronic and diachronic perspective.
It seems that reflexes of the detransitivizer \rc{ət(e)-} were optionally added to an \gl{s_a_} verb root \rc{jəpɨ}, which further underwent umlaut and loss of \rc{j}, but in no systematic manner, resulting in the chaotic picture in \cref{tab:come}.

As discussed in \cref{sec:pekodian}, innovative \rc{k-} was introduced on the \arara reflex of \rc{jəpɨ}, but not on the \ikpeng and \bakairi reflexes of \rc{ətjəpɨ}.
Both reflexes of \rc{ətjəpɨ} (\trio) and of \rc{ətjəpɨ} and/or \rc{epɨ} (\akuriyo) resisted the introduction of \rc{t-} in \PTir.
\carijo \obj{ehɨ} shows innovative \obj{j-}, rather than conservative \obj{w-} \exref{car-18}.
It is unknown whether there is a \yukpa reflex of this verb, and it was fully replaced in \PWai by \rc{omokɨ} \qu{to come}.

@@

\begin{table}
	\centering
	\caption{Reflexes of \rc{ɨpɨtə} \qu{to go down} \parencites[294, 325, 331]{triomeira1999}[196]{hixkaryanaderby1979}[55]{waiwaihawkins1998}[118]{guerrero2019carijo}[44]{camargo2010wayana}[99]{camargo2002lexico}[263]{courtz2008carib}[450]{maquiritaricaceres2011}[139]{stegeman2014akawaio}[139]{alvarez2008clausulas}[34]{macushiabbott1991}[88]{mattei1994diccionario}[68]{mendez1959yawarana}[58]{bruno1996dictionary}[84]{gildea1994akuriyo}[153]{alves2017arara}[137]{von1892bakairi}[116; p.c., Angela Fabíola Alves Chagas, David Felipe Guerrero, Spike Gildea]{meira1998proto}}
	\label{tab:XXX}
\begin{tabular}[t]{@{}lll@{}}
\toprule
Language &            Form &      Class \\
\midrule
\PPar     &     \obj{ɨɸɨto} &  \gl{s_p_} \\
\kaxui    &     \obj{ɨhɨto} &  \gl{s_p_} \\
\hixka    &       \obj{hto} &          ? \\
\waiwai   &       \obj{hto} &          – \\
\PPek     &      \obj{ɨptə} &  \gl{s_a_} \\
\arara    &     \obj{iptoŋ} &  \gl{s_a_} \\
\ikpeng   &     \obj{iptoŋ} &          ? \\
\bakairi  &     \obj{ɨtəgɨ} &  \gl{s_a_} \\
\PTir     &      \obj{ɨhtə} &  \gl{s_a_} \\
\trio     &      \obj{ɨhtə} &  \gl{s_a_} \\
\akuriyo  &       \obj{ɨtə} &  \gl{s_a_} \\
\carijo   &     \obj{ehɨtə} &          – \\
\wayana   &      \obj{ɨptə} &  \gl{s_p_} \\
\apalai   &      \obj{ɨhto} &  \gl{s_p_} \\
\kalina   &  (\obj{onɨʔto}) &  \gl{s_a_} \\
\maqui    &      \obj{əʔtə} &  \gl{s_p_} \\
\kapon    &    (\obj{uʔtə}) &          – \\
\pemon    &    (\obj{uʔtə}) &          – \\
\macushi  &    (\obj{autɨ}) &          – \\
\panare   &      \obj{əhtə} &  \gl{s_a_} \\
\yawarana &      \obj{əhtə} &          – \\
\waimiri  &       \obj{ɨtɨ} &          – \\
\bottomrule
\end{tabular}
\end{table}
